Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-02-10 LaTeX/Turnover.tex
references

%CHECK THROUGHOUT: use of sex-determination SYSTEMS versus MECHANISMS.

@@

It has been suggested that sex ratio selection could be a particularly important force driving transitions between sex-determining systems \citep[Chapter 7]{Beukeboom:2014vb}. 
The default mode of sex ratio selection is `Fisherian' sex ratio selection, which favours equal investment in male and female offspring (i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce)  \textcolor{red}{(Fisher)}. 
Given that the sex determination system can directly affect the sex ratio, we might expect Fisherian sex ratio selection to influence the spread of new sex determination systems. 
For example, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, which restoring an even sex ratio.

We note that non-Fisherian (`extraordinary') sex ratios can be favoured in some situations. 
For example, with local mate competition, where brothers compete for mates, a female biased sex ratio is favoured \citep{Hamilton:1967ts}. 
Thus, local mate competition could favour new sex determination systems because feminizing mutations bias the sex ratio towards females \citep{Wilson:1981vm,Vuillleumier:2007bh}. 
Another consideration is that environmental conditions (e.g., maternal condition, mate quality, age, or host size) can differentially affect the fitness of males versus females such that the optimal allocation to males/females depends on the environment (Trivers and Willard, 1973) \citep{Charnov:1977tx} (Charnov, 1982, West 2009, Chapters 6 and 7). 
In such cases, flexible sex determination systems may evolve in order to allow the primary sex ratio to be determined in a way that depends on the environment \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In this study, we do not consider environmental condition dependence or local mate competition (reviewed in Charnov, 1982 and West 2009).

%\textcolor{red}{We add haploid selection:}
Here, we use mathematical models to find the conditions under which new sex determination systems are favoured when loci experience haploid selection. 
Haploid genotypes at many loci experience selection during gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-medelian) segregation of genotypes during gamete production and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool; the term `haploid selection' encompasses both processes. 
Meiotic drive generally occurs either during the production of male or female gametes only \textcolor{red}{(Ubeda and Haig, 2005; Lindholm et al. 2016)}. 
Because there are typically more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.

@@

Under Fisherian sex ratio selection, autosomal modifiers favour equal investment in male and female offspring, i.e., a 1:1 sex ratio \textcolor{red}{(Fisher... others)}. 
Therefore, we are particularly interested in the case where the novel environmental sex-determiner causes half of its carriers to become female and half to become male ($k=1/2$). 
As with an autosomal locus, these sex-determiners will be found in males half of the time and in females half of the time.
However, we find that the growth rate of a rare, dominant offspring-controlled neo-ESD allele that produces males or females with equal probability ($k=1/2$) is
\begin{equation}
\lambda_{ESD',XY} =1+ \frac{1}{2}\frac{(\lambda_{Y',XY}-1) + (\lambda_{W',XY}-1)}{2} \Big|_{R=1/2} + O\left(\epsilon^3\right),
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s^\male - 3 s^\female\right) }{8 \left( s_{f} + s_{m} \right)^2} t_{m}^2.
\label{eq:lambda_ESD}
\end{equation}
\noindent
We note that recombination between the selected locus and the novel sex-determining locus, $R$, doesn't enter into the $k=1/2$ results because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-$W$ and half the time as a neo-$Y$.
preventing associations from building up between allele $A$ and sex.  
Equation \eqref{eq:lambda_ESD} shows that invasion by a novel `perfect' ESD (equal sex ratio, $k=1/2$) mutation is the same for an ancestrally XY or ZW system (since $\lambda_{Y',XY} = \lambda_{W',ZW}$, $\lambda_{W',XY} = \lambda_{Y',ZW}$).
Thus, by the same argument as above (if drive occurs in males only then the sex ratio is only biased when the ancestral sex-determination system is XY), Fisherian sex ratio selection does not explain invasion by an offspring-controlled neo-ESD locus. 
Rather, the neo-ESD gets half of the fitness of a feminizing mutation (neo-$W$) and half of the fitness of a masculizing mutation (neo-$Y$), but only has an effect one half of the time (the other half of the time it produces the same sex as the ancestral system would have, to leading order). 
The net result can be that perfect ESD will not invade, even if sex ratios are biased. 
For example, if there is haploid selection in males (either drive or pollen/sperm competition) but the conditions in table \ref{tab:specialcases} are not met, perfect ESD will not invade, even though it would equalize the zygotic sex ratio from an initially biased case (assuming $r<1/2$).

@@

One might expect Fisherian sex ratio selection to influence the spread of new sex-determining systems in our models because linkage between haploid selected loci and sex-determining regions causes biased zygotic sex ratios \citep{Hamilton:1967ts,Burt:2006,Field:2012fd,Field:2013cc}. 
Fisherian sex ratio selection follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male, and half from a female. 
Thus, if the population sex ratio is biased towards females, the average per-individual contribution of genetic material to the next generation from males is greater than the contribution from females (and vice versa for male-biased sex ratios). 
Therefore, a mutant that increases investment in males will spread via the higher per-individual contributions made by males. 
%That is, under Fisherian sex ratio selection, the success of a mutant relative to the non-mutant depends, in equal parts, on the contributions made by males and females to the next generation. 
An implicit assumption of Fisherian sex ratio selection is that the mutant allele is autosomal and has the same inheritance pattern as the non-mutant allele. 
The mutations we consider here, neo-sex-determining alleles, break this assumption. 
For example, the success of neo-Y/neo-W mutations depends only on the number of alleles contributed by males/females (Table \ref{tab:haplotype_growth}). 
In this respect, a neo-W is similar to a cytoplasmic element, which also do not experience selection to balance sex ratios \textcolor{red}{citations}.
Even mutants that are equally likely to be found in males or females, such as an environmental sex determination mutation (equation \ref{eq:lambda_ESD}), are not strictly autosomal if they determine sex. 
Thus, despite the fact that sex ratio biases caused by gametic competition or meiotic drive have been shown to exert Fisherian sex ratio selection on various autosomal modifiers \citep{Stalker:1961th,Smith:1975ft,Frank:1989vl,Hough:2013uo,Ubeda:2015fx, Otto:2015va}, we do not find evidence of Fisherian sex ratio selection acting during invasion by neo-sex-determination systems (e.g., see Figure \ref{fig:Combination_Turnover} and \citealt{Ubeda:2015fx}, in which a neo-Y invades despite biasing sex ratios).

@@

In addition, our model of meiotic drive is simple, involving a single locus with two alleles. 
However, many meiotic drive systems involve an interaction with another locus at which alleles may `suppress' the action of meiotic drive \citep{Burt:2006} \textcolor{red}{(Lindholm et al. 2016)}. 
Thus, the dynamics of meiotic drive alleles can be heavily dependent on the interaction between two loci and the recombination rate between them, which in turn can be affected by sex-linkage if there is reduced recombination between sex chromosomes \textcolor{red}{(Hurst and Pomiankowski, 1991)}. 
Furthermore, in some cases, a driving allele may act by killing any gametes that carry a `target' allele at another locus, in which case there is a two-locus drive system and the total number of gametes produced can be reduced by meiotic drive. 
Where gamete number is reduced by meiotic drive, the number of mates competing for fertilization (mating system) can affect the equilibrium frequency of a meiotic drive allele \textcolor{red}{Holman et al., 2015)}. 
In polygamous mating systems, the intensity of pollen/sperm competition can depend on the density of males available to donate pollen/sperm, which can itself depend on the sex ratio \textcolor{red}{(Taylor and Jaenike, 2002)}. 
Since the sex ratio is partly determined by the sex determination system, the evolution of new sex determination system could by influenced by these dynamics. 
It remains to be investigated how the evolution of new sex-determining mechanisms could be influenced by two-locus meiotic drive and by ecological feedbacks under different mating systems.

The hypotheses presented here can be investigated in a similar manner to the idea that transitions between sex-determining systems are favoured by linkage to sexually antagonistic variation. 
In the case of sexually antagonistic variation, one supporting observation is that genes that appear to experience sexually-antagonistic selection have been found on recently derived sex chromosomes \textcolor{red}{CHECK (Kallman 1973; Wada et al. 1998; Lande et al. 2001; Lindholm and Breden 2002; Streelman et al. 2003; Fernandez and Morris 2008; Kitano et al. 2009)} \textcolor{red}{(Roberts et al. 2009; Ser et al., 2010; Kitano,2012)}.
However, it is possible that sexually antagonistic variation accumulated after sex chromosome transitions \textcolor{red}{(CHECK: Ser et al., 2010)} because linkage with the sex-determining regions allows sexually antagonistic selection to maintain polymorphisms under a larger parameter space \textcolor{red}{(Rice, 1987, Jordan and Charlesworth, 2010-ish)}.
We note that linkage with sex chromosomes is not, a priori, more permissive to the maintainence of ploidally antagonistic variation \citep{Immler:2012tl}.
Secondly, we have shown that new sex-determination systems can be favoured if either the ancestral sex-determining region or the new sex-determining region are linked to loci under haploid selection. 
Therefore, the presence of haploid selected loci around ancestral- or new- sex-determining regions could support their role in sex chromosome turnover. 
As with sexually-antagonistic variation, a comparison between closely related clades could indicate whether a polymorphism pre-dates a transition in sex-determination or arose afterwards. 
%It would be particularly informative if selection were evaluated in closely related clades

Taken at face value, our results indicate that transitions in heterogametey (XY to ZW or vice versa) are more likely to be favoured by selection if there is selection upon both haploid and diploid genotypes rather than diploid selection alone. 
This prediction could be examined using a suitable proxy for haploid selection, for example, \textcolor{red}{Lenormand and Dutheil (2005)} use the outcrossing rate as a proxy for the strength of pollen competition. 
In animals, one might expect gametic competition to be stronger in species where sperm is required to live for a long time after spermatogenesis because transcripts shared during spermatogenesis may become depleted, revealing the haploid phenotype of the sperm \citep{Immler:2014im}. 
Given the caveats mentioned above about the form of meiotic drive modelled, we would also expect that heterogametic transitions in sex determination would be more common in clades where there is meiotic drive.

We have shown that haploid selection can drive transitions between sex determination systems, such that haploid selection is incorporated in the factors influencing the evolution of sex determination. 
However, the particular way in which transitions are affected by haploid selection is not intuitively obvious. 
Firstly, sex-specific haploid selection affects turnovers between sex determination systems in a manner that is qualitatively different from diploid sex-specific selection. 
In particular, closer linkage between a sex-determining locus and a selected locus is not always favoured during heterogametic transitions when there is haploid selection. 
Secondly, even though haploid selection is a source of zygotic sex ratio biases, Fisherian sex ratio selection does not have good explanatory power in our models in determining whether various sex-determination systems evolve; this result is surprising given that sex ratios are ultimately determined via the sex-determination system.

>>>>

%CHECK THROUGHOUT: use of sex-determination SYSTEMS versus MECHANISMS. 
%CHECK THROUGHOUT: use of haploid selection, gametic selection, and meiotic drive.

@@

It has been suggested that sex ratio selection could be a particularly important force driving transitions between sex-determining systems \citep[Chapter 7]{Beukeboom:2014vb}. 
The default mode of sex ratio selection is `Fisherian' sex ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
Given that the sex determination system can directly affect the sex ratio, we might expect Fisherian sex ratio selection to influence the spread of new sex determination systems. 
For example, \citet{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased \citep{Hamilton:1967ts}. 
They find that new, unlinked sex-determining loci (masculinizing or feminizing mutations, i.e., neo-Y or neo-W loci) can then spread, which restoring an even sex ratio.

We note two other ways in which sex determination has been shown to relate to zygotic sex ratios.
Firstly, non-Fisherian female-biased sex ratios can be favoured when there is local mate competition, in which brothers compete for mates \citep{Hamilton:1967ts}. 
Therefore, with local mate competition, feminizing mutations can spread because they bias the sex ratio towards females \citep{Wilson:1981vm,Vuillleumier:2007bh}. 
Secondly, environmental conditions (e.g., maternal condition, mate quality, age, or host size) can differentially affect the fitness of males versus females such that the optimal allocation to males/females depends on the environment \citep{Trivers:1973wb,Charnov:1977tx,Charnov:1982wg}. 
In such cases, flexible sex determination systems may evolve in order to allow the zygotic sex ratio to be determined in a way that depends on the environment \citep{Charnov:1977tx,Werren:1984tl,Pen:2010kk}. 
In this study, we do not consider environmental condition dependence or local mate competition \citep[reviewed in][]{Charnov:1982wg,West:2009we}.

%\textcolor{red}{We add haploid selection:}
Here, we use mathematical models to find the conditions under which new sex determination systems are favoured when loci experience haploid selection. 
Haploid genotypes at many loci experience selection during gamete competition and/or meiotic drive \citep{Mulcahy:1996ha,JOSEPH:2004haa}.
We use the term `meiotic drive' to refer to the biased (non-medelian) segregation of genotypes during gamete production and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool; the term `haploid selection' encompasses both processes. 
Meiotic drive generally occurs either during the production of male or female gametes only \citep{Ubeda:2005gw,Lindholm:2016cw}.
Because there are typically more pollen/sperm than required for fertilization, gametic competition is also typically sex specific, occurring primarily among male gametes.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte and these genes exhibit stronger signatures of selection than random genes \citep{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes cause the frequency of resistant alleles to increase \citep[e.g.,][]{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}. 
A smaller (but non-negligible) proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain \citep{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}.

@@

Under Fisherian sex ratio selection, autosomal modifiers favour equal investment in male and female offspring, i.e., a 1:1 sex ratio \citep{Fisher:1930wy,Charnov:1982wg,West:2009we}. 
Therefore, we are particularly interested in the case where the novel environmental sex-determiner causes half of its carriers to become female and half to become male ($k=1/2$). 
As with an autosomal locus, these sex-determiners will be found in males half of the time and in females half of the time.
However, we find that the growth rate of a rare, dominant offspring-controlled neo-ESD allele that produces males or females with equal probability ($k=1/2$) is
\begin{equation}
\lambda_{ESD',XY} =1+ \frac{1}{2}\frac{(\lambda_{Y',XY}-1) + (\lambda_{W',XY}-1)}{2} \Big|_{R=1/2} + O\left(\epsilon^3\right),
%V_A \frac{r}{1-2r}  \frac{s_{f} \left( s^\male - 3 s^\female\right) }{8 \left( s_{f} + s_{m} \right)^2} t_{m}^2.
\label{eq:lambda_ESD}
\end{equation}
\noindent
We note that recombination between the selected locus and the novel sex-determining locus, $R$, doesn't enter into the $k=1/2$ results because sex is essentially randomized each generation, %as if the novel sex-determiner were on an unlinked site that half the time acts as a neo-$W$ and half the time as a neo-$Y$.
preventing associations from building up between allele $A$ and sex.  
Equation \eqref{eq:lambda_ESD} shows that invasion by a novel `perfect' ESD (equal sex ratio, $k=1/2$) mutation is the same for an ancestrally XY or ZW system (since $\lambda_{Y',XY} = \lambda_{W',ZW}$, $\lambda_{W',XY} = \lambda_{Y',ZW}$).
Thus, by the same argument as above (if drive occurs in males only then the sex ratio is only biased when the ancestral sex-determination system is XY), Fisherian sex ratio selection does not explain invasion by an offspring-controlled neo-ESD locus. 
Rather, the neo-ESD gets half of the fitness of a feminizing mutation (neo-$W$) and half of the fitness of a masculizing mutation (neo-$Y$), but only has an effect one half of the time (the other half of the time it produces the same sex as the ancestral system would have, to leading order). 
The net result can be that perfect ESD will not invade, even if sex ratios are biased. 
For example, if there is haploid selection in males (either drive or pollen/sperm competition) but the conditions in table \ref{tab:specialcases} are not met, perfect ESD will not invade, even though it would equalize the zygotic sex ratio from an initially biased case (assuming $r<1/2$).

@@

One might expect Fisherian sex ratio selection to influence the spread of new sex-determining systems in our models because linkage between haploid selected loci and sex-determining regions causes biased zygotic sex ratios \citep{Hamilton:1967ts,Burt:2006,Field:2012fd,Field:2013cc}. 
Fisherian sex ratio selection follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male, and half from a female \citep{Fisher:1930wy,West:2009we}. 
Thus, if the population sex ratio is biased towards females, the average per-individual contribution of genetic material to the next generation from males is greater than the contribution from females (and vice versa for male-biased sex ratios). 
Therefore, a mutant that increases investment in males will spread via the higher per-individual contributions made by males. 
%That is, under Fisherian sex ratio selection, the success of a mutant relative to the non-mutant depends, in equal parts, on the contributions made by males and females to the next generation. 
An implicit assumption of Fisherian sex ratio selection is that the mutant allele is autosomal and has the same inheritance pattern as the non-mutant allele. 
The mutations we consider here, neo-sex-determining alleles, break this assumption. 
For example, the success of neo-Y/neo-W mutations depends only on the number of alleles contributed by males/females (Table \ref{tab:haplotype_growth}). 
In this respect, a neo-W is similar to a cytoplasmic element, which also do not experience selection to balance sex ratios \citep{Frank:1989vl,Werren:1998co,Chase:2007bf}.
Even mutants that are equally likely to be found in males or females, such as an environmental sex determination mutation (equation \ref{eq:lambda_ESD}), are not strictly autosomal if they determine sex. 
Thus, despite the fact that sex ratio biases caused by gametic competition or meiotic drive have been shown to exert Fisherian sex ratio selection on various autosomal modifiers \citep{Stalker:1961th,Smith:1975ft,Frank:1989vl,Hough:2013uo,Ubeda:2015fx, Otto:2015va}, we do not find evidence of Fisherian sex ratio selection acting during invasion by neo-sex-determination systems (e.g., see Figure \ref{fig:Combination_Turnover} and \citealt{Ubeda:2015fx}, in which a neo-Y invades despite biasing sex ratios).

@@

In addition, our model of meiotic drive is simple, involving a single locus with two alleles. 
However, many meiotic drive systems involve an interaction with another locus at which alleles may `suppress' the action of meiotic drive \citep{Burt:2006,Lindholm:2016cw}.
Thus, the dynamics of meiotic drive alleles can be heavily dependent on the interaction between two loci and the recombination rate between them, which in turn can be affected by sex-linkage if there is reduced recombination between sex chromosomes \citep{Hurst:1991uh}.
Furthermore, in some cases, a driving allele may act by killing any gametes that carry a `target' allele at another locus, in which case there is a two-locus drive system and the total number of gametes produced can be reduced by meiotic drive. 
Where gamete number is reduced by meiotic drive, the number of mates competing for fertilization (mating system) can affect the equilibrium frequency of a meiotic drive allele \citep{Holman:2015en}. 
In polygamous mating systems, the intensity of pollen/sperm competition can depend on the density of males available to donate pollen/sperm, which can itself depend on the sex ratio \citep{Taylor:2002wu}. 
Since the sex ratio is partly determined by the sex determination system, the evolution of new sex determination system could by influenced by these dynamics. 
How the evolution of new sex-determining mechanisms could be influenced by two-locus meiotic drive and/or by ecological feedbacks under different mating systems remains to be studied.

The hypotheses presented here can be investigated in a similar manner to the idea that transitions between sex-determining systems are favoured by linkage to sexually antagonistic variation. 
In the case of sexually antagonistic variation, one supporting observation is that genes that appear to experience sexually-antagonistic selection have been found on recently derived sex chromosomes \citep{Lindholm:2002dw,Tripathi:2009cw,Ser:2010iq}.
However, it is possible that sexually antagonistic variation accumulated after sex chromosome transitions because linkage with the sex-determining regions allows sexually antagonistic selection to maintain polymorphisms under a larger parameter space \citep{Rice:1987hs,Jordan:2011fj}.
We note that linkage with sex chromosomes is not, a priori, more permissive to the maintainence of ploidally antagonistic variation \citep{Immler:2012tl}.
However, as with sexually-antagonistic variation, a comparison between closely related clades could indicate whether a polymorphism pre-dates a transition in sex-determination or arose afterwards. 
Secondly, we have shown that new sex-determination systems can be favoured if either the ancestral sex-determining region or the new sex-determining region are linked to loci under haploid selection. 
Therefore, the presence of haploid selected loci around ancestral- or new- sex-determining regions could support their role in sex chromosome turnover. 
%It would be particularly informative if selection were evaluated in closely related clades

Taken at face value, our results indicate that transitions in heterogametey (XY to ZW or vice versa) are more likely to be favoured by selection if there is selection upon both haploid and diploid genotypes rather than diploid selection alone. 
This prediction could be examined using a suitable proxy for haploid selection, for example, \citep{Lenormand:2005vb} use the outcrossing rate as a proxy for the strength of pollen competition. 
In animals, one might expect gametic competition to be stronger in species where sperm is required to live for a long time after spermatogenesis because transcripts shared during spermatogenesis may become depleted, revealing the haploid phenotype of the sperm \citep{Immler:2014im}. 
Given the caveats mentioned above about the form of meiotic drive modelled, we would also expect that heterogametic transitions in sex determination would be more common in clades where there is meiotic drive.

We have shown that haploid selection can drive transitions between sex determination systems, such that haploid selection is should be incorporated into the factors that influence the evolution of sex determination. 
However, the particular way in which transitions are affected by haploid selection is not intuitively obvious. 
Firstly, sex-specific haploid selection affects turnovers between sex determination systems in a manner that is qualitatively different from diploid sex-specific selection. 
In particular, closer linkage between a sex-determining locus and a selected locus is not always favoured during heterogametic transitions when there is haploid selection. 
Secondly, even though haploid selection is a source of zygotic sex ratio biases, Fisherian sex ratio selection does not have good explanatory power in our models in determining whether various sex-determination systems evolve; this result is surprising given that sex ratios are ultimately determined via the sex-determination system.
